Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-03 popl-rebuttal.md

We thank the reviewers for their valuable feedback.

Reviewer A
-----------

- Read Uncommitted (RU) isolation: While RU is a valid ANSI SQL
  isolation level, it is typically not implemented by modern
  databases. For example, Oracle supports only RC and SER, while
  Postgres maps RU to RC. Even when RU is supported, it is never the
  default isolation level (Bailis et al, HotOS'13)., and applications
  rarely demote isolation levels to RU in practice (Bailis et al,
  SIGMOD'15). Furthermore, RU violates atomicity, the basic tenet
  underlying transactions.  Given its lack of acceptance in the
  database community, and the complications it would introduce to our
  formalization, we opted to use RC as the baseline isolation level.

- Limited set of operations in the DSL: While it is true our language
  only includes four basic SQL operations, they have proven sufficient
  to implement widely accepted benchmarks, such TPC-C and TPC-E. We
  also note that MySQL's [developer
  documentation](https://dev.mysql.com/doc/internals/en/) (also see: S
  Pachev, Understanding MySQL Internals, 1st Edition, Chapter 9)
  indicates that its core calculus for SQL is indeed composed of these
  four basic operations, making our technique likely applicable on
  MySQL's core language, if not its surface language. We expect
  similar conclusions to hold on other databases.

Reviewer B
-------------

- Isolation levels weaker than RC: please see above.

- Not admitting valid snapshot isolation (SI) executions:  We believe
  the reviewer has misunderstood our specification for SI as $I_{ss}$,
  which only describes the semantics of "executing against a
  snapshot". SI indeed executes against a snapshot (thus its $I_e =
  I_{ss}$), but it *commits* against any non-write-write-conflicting
  database state (thus its $I_c = I_{ww}$). This is explained on p.13
  ("Snapshot Isolation"). The presented specification indeed allows
  non-conflicting concurrent transactions to commit in parallel, and hence
  admits all SI executions.

- Decidability: The RG-TXN (Sec. 5.2, p. 20) rule does not lead to
  second-order quantification or quantified formulas with negation at
  prenex position. Sets are encoded as uninterpreted functions (Fig.
  10), and quantified set variables are simply replaced with new
  function symbols that occur free in the encoded formula.  Consider
  the ${\sf F} \Rightarrow G$ premise of RG-TXN. The quantified set
  variables $\delta$ and $\Delta$ are instantiated with fresh function
  variables $f_{\delta}$ and $f_{\Delta}$, respectively, resulting in
  a quantifier-free implication with free function symbols. The
  antecedent of this implication (${\sf F}$) is asserted as such, but
  its consequent ($G$) is negated before checking the satisfiability
  of the entire formula. This is a standard approach often employed to
  discharge verification conditions of the form $\forall a. \phi_1(a)
  \Rightarrow \phi_2(a)$.  In this process: (i). sets or functions are
  never quantified (thus no second-order quantification), and
  (ii). ${\sf F}$ is never negated (thus no inversion of quantifiers
  in ${\sf F}$). Other premises are similarly encoded.

  The reviewer is correct in observing that our encoding is not in the
  EPR fragment of FOL.  However, our encoding is still decidable as it
  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
  that admits sentences of the form:
  $\exists^{*}\forall^{2}\exists^{*}$ (see: Borger et al, The
  Classical Decision Problem, 1st Edition, p.261, and Gurevich, On the
  Classical Decision Problem, p.7); formulas generated by our encoding
  clearly belong to this fragment.  We will clarify this issue in the
  final version.

- Local Context Independence: This assumption is used by the inference
  rule for `FOREACH`. The rule computes the transformer of a loop by
  lifting the transformer (${\sf F}$) of its body. In the process, it
  assumes that the effect of each loop iteration depends only on
  the loop variable $z$, but not on local ($\delta$) and global
  ($\Delta$) states (observe that ${\sf F}$ is applied to the same
  $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
  gauranteed by the stability condition (checked separately), the
  independence on the local state $\delta$ is simply assumed.

Reviewer C
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

### 2017-10-03 popl-rebuttal.txt

We thank the reviewers for their valuable feedback.

Reviewer A
-----------

- Read Uncommitted (RU) isolation:  While RU is a valid ANSI SQL
  isolation level, it is typically not implemented by modern
  databases. For example, Oracle supports only RC and SER, while
  Postgres maps RU to RC. Even when RU is supported, it is never the
  default isolation level (Bailis et al, HotOS'13)., and applications
  rarely demote isolation levels to RU in practice (Bailis et al,
  SIGMOD'15). Thus, RC can be considered as a baseline isolation level
  that meets the performance demands of real-world applications.
  Furthermore, RU violates atomicity, the basic tenet underlying
  transactions.  Given its lack of acceptance in the database
  community, and the complications it would introduce to our
  formalization, we opted to elide it from the development.

- Limited set of operations in the DSL:  While it is true our language
  only includes four basic SQL operations, they have proven sufficient
  to implement widely accewpted standard benchmarks, such TPC-C and
  TPC-E. We also note that MySQL's [developer
  documentation](https://dev.mysql.com/doc/internals/en/) (also see: S
  Pachev, Understanding MySQL Internals, 1st Edition, Chapter 9)
  indicates that its core calculus for SQL is indeed composed of these
  four basic operations, making our technique likely applicable on
  MySQL's core language, if not its surface language. We expect
  similar conclusions to hold on other databases.

@@

Reviewer B
-------------

- Isolation levels weaker than RC: please refer to our comments on RU
  isolation above.

- Not admitting valid snapshot isolation (SI) executions:  We believe
  the reviewer has misunderstood our specification for SI as $I_{ss}$,
  which only describes the semantics of "executing against a
  snapshot". SI indeed executes against a snapshot (thus its $I_e =
  I_{ss}$), but it *commits* against any non-write-write-conflicting
  database state (thus its $I_c = I_{ww}$). This is explained on p.13
  ("Snapshot Isolation"). The presented specification indeed allows
  non-conflicting concurrent transactions to commit in parallel, hence
  admits all SI executions.

- Decidability: The RG-TXN (Sec. 5.2, p. 20) rule does not lead to
  second-order quantification or quantified formulas with negation at
  prenex position. Sets are encoded as uninterpreted functions (Fig.
  10), and quantified set variables are simply replaced with new
  function symbols that occur free in the encoded formula. For
  instance consider the ${\sf F} \Rightarrow G$ premise of RG-TXN. The
  quantified set variables $\delta$ and $\Delta$ are instantiated with
  fresh function variables $f_{\delta}$ and $f_{\Delta}$,
  respectively, resulting in a quantifier-free implication with free
  function symbols. The antecedent of this implication (${\sf F}$) is
  asserted as such, but its consequent ($G$) is negated before
  checking the satisfiability of the entire formula. This is a
  standard approach often employed to discharge the verification
  conditions of form $\forall a. \phi_1(a) \Rightarrow \phi_2(a)$.
  Note that in this process: (i). sets or functions are never
  quantified (thus no second-order quantification), and (ii). ${\sf
  F}$ is never negated (thus no inversion of quantifiers in ${\sf
  F}$). Other premises are similarly encoded.

  The reviewer is correct in observing that our encoding is not in the
  EPR fragment of FOL since we admit formulas of the form $\forall
  x.\exists y.\phi$.  However, our encoding is still decidable as it
  belongs to a closely related variant of EPR, Godel-Kalmar-Schutte,
  that admits sentences of the form:
  $\exists^{*}\forall^{2}\exists^{*}$ (see: Borger et al, The
  Classical Decision Problem, 1st Edition, p.261, and Gurevich, On the
  Classical Decision Problem, p.7); formulas generated by our encoding
  clearly belong to this fragment.  We will clarify this issue in the
  final version.

### 2017-10-10 motivation.tex

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, such a conflict occurs if
concurrentlyh executing transactions modify the same record) among
concurrent transactions when they commit. Write-write conflicts can be
eliminated in various ways, e.g., through conflict detection followed
by a rollback, or through exclusive locks, or a combination of both.
For instance, one possible implementation of SI, close to the one used
by PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion indicates, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction are not visible to another concurrently executing one
(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witness
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the tuples that match the search
criteria from $\Delta$ and computes the updated tuples that are to be
added to the local database. Thus, the state transformer for the update
statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
The advantage of inferring such transformers is that we can now make
use of a semantics-preserving translation from the domain of sets
equipped with $\bind$ to a decidable fragment of first-order logic,
allowing us to leverage SMT solvers for automated proofs without
having to infer potentially complex thread-local invariants or
intermediate assertions.  Sec.~\ref{sec:inference} describes this
translation. In the exposition thus far, we assumed $\Delta$ remains
invariant, which is clearly not the case when we admit concurrency.
Necessary concurrency extensions of the state transformer semantics to
deal with interference is also covered in Sec.~\ref{sec:inference}.
The following two sections, however, first focus on laying the
theoretical foundations for reasoning about weak isolation.

>>>>

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundness of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, such a conflict occurs if
concurrently executing transactions modify the same record) among
concurrent transactions when they commit. Write-write conflicts can be
eliminated in various ways, e.g., through conflict detection followed
by a rollback, or through exclusive locks, or a combination of both.
For instance, one possible implementation of SI, close to the one used
by PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion indicates, implementations of SI on real databases
such as PostgreSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction are not visible to another concurrently executing one
(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witness
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the tuples that match the search
criteria from $\Delta$ and computes the updated tuples that are to be
added to the local database. Thus, the state transformer for the update
statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becomes set monadic $\bind$.
The advantage of inferring such transformers is that we can now make
use of a semantics-preserving translation from the domain of sets
equipped with $\bind$ to a decidable fragment of first-order logic,
allowing us to leverage SMT solvers for automated proofs without
having to infer potentially complex thread-local invariants or
intermediate assertions.  Sec.~\ref{sec:inference} describes this
translation. In the exposition thus far, we assumed $\Delta$ remains
invariant, which is clearly not the case when we admit concurrency.
Necessary concurrency extensions of the state transformer semantics to
deal with interference is also covered in Sec.~\ref{sec:inference}.
The following two sections, however, first focus on laying the
theoretical foundations for reasoning about weak isolation.

### 2017-10-10 reasoning.tex

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It depends on a
transaction-local judgment to verify the body ($c$) of a transaction with
id $i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_e$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
\noindent While $R$ describes changes in the global database state, $\I_e$
constrains the global effects visible to the transaction.
Recall however that $\I_e$ uses the global state in effect at the
point the transaction last took a step ($\stg_1$) to determine whether
the transaction can proceed from the current global state ($\stg$).
Moreover, the global rely relation $R$ defines the permissible
interferences from $\Delta$ to a future global state ($\Delta'$).  The
local rely relation $\R_e$ determines allowed interferences as
specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
$\stg_1$ defines this reley relation to be an over-approximation on
the interferences in the presence of which the transaction can
proceed, simplifying verification burden.

>>>>

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It depends on a
transaction-local judgment to verify the body ($c$) of a transaction with
id $i$. The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_e$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_e(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_e(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
\noindent While $R$ describes changes in the global database state, $\I_e$
constrains the global effects visible to the transaction.
Recall however that $\I_e$ uses the global state in effect at the
point the transaction last took a step ($\stg_1$) to determine whether
the transaction can proceed from the current global state ($\stg$).
Moreover, the global rely relation $R$ defines the permissible
interferences from $\Delta$ to a future global state ($\Delta'$).  The
local rely relation $\R_e$ determines allowed interferences as
specified by $\I_e$ assuming $\stg_1$.  In effect, the existential on
$\stg_1$ defines this rely relation to be an over-approximation on
the interferences in the presence of which the transaction can
proceed, simplifying verification burden.

### 2017-10-10 related-work.tex

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, and the kinds of
dependencies that are forbidden in each case. The operational nature
of Adya's specifications make them suitable for runtime monitoring and
anomaly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
the declarative nature of our specifications make them suitable for
formal reasoning about program behaviour. Sivaramakrishnan \emph{et
al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, making it difficult
to reason about a program as it builds the trace. Cerone \emph{et
al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
visibility, but their specifications are also for complete traces.
Like ours, all the aforementioned specification frameworks use the
vocabulary introduced in~\cite{burckhardt14}. However, none of them
are equipped with a reasoning framework that can use such
specifications to verify programs under weak isolation.

@@

\vspace*{-4pt}
\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the context of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar (automated inference of intermediate
assertions and local invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), the properties being
verified (integrity constraints over relational tables vs. memory
safety, or linearizability of concurrent data strucgture operations)
and the analysis technique used to drive inference (state transformers
vs. abstract interpretation) are quite different.

>>>>

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, and the kinds of
dependencies that are forbidden in each case. The operational nature
of Adya's specifications make them suitable for runtime monitoring and
anomaly detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas
the declarative nature of our specifications make them suitable for
formal reasoning about program behavior. Sivaramakrishnan \emph{et
al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, making it difficult
to reason about a program as it builds the trace. Cerone \emph{et
al.}~\cite{gotsmanconcur15} specify isolation levels with atomic
visibility, but their specifications are also for complete traces.
Like ours, all the aforementioned specification frameworks use the
vocabulary introduced in~\cite{burckhardt14}. However, none of them
are equipped with a reasoning framework that can use such
specifications to verify programs under weak isolation.

@@

\vspace*{-4pt}
\paragraph{Inference.}  Prior work by Vafeiadis~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the context of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar (automated inference of intermediate
assertions and local invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), the properties being
verified (integrity constraints over relational tables vs. memory
safety, or linearizability of concurrent data structure operations)
and the analysis technique used to drive inference (state transformers
vs. abstract interpretation) are quite different.
